---
title: "Akula-DATA607-Week03-Project01"
author: "Pavan Akula"
date: "February 21, 2017"
output: html_document
---

This project involves in reading *_Chess Tournament Cross Table_* and generating a CSV file with the following information for all of the players: Player's Name, Player's State, Total Number of Points, Player's Pre-Rating, and Average Pre Chess Rating of Opponents.

Average Pre Chess Rating of Opponents is calculated as sum of pre-tournament opponents' ratings divided by total number of games played.

Libraries used for this project.

```{r , echo=TRUE}
library(readr)      # used for reading the text file
library(stringr)    # used for extracting text using regular expressions
library(data.table) # used for saving out into tabel format
library(sqldf)      # used for querying and to generate aggregate output
library(utils)      # used for generating CSV file
```

Download and save the tournamentinfo.txt, to local working directory from website.
https://bbhosted.cuny.edu/bbcswebdav/pid-28524631-dt-content-rid-126144833_1/courses/SPS01_DATA_607_01_1172_1/SPS01_DATA_607_01_1169_1_ImportedContent_20160815114002/SPS01_DATA_607_01_1162_1_ImportedContent_20160126062155/tournamentinfo.txt

Load *_Chess Tournament Cross Table_* file. 

```{r , echo=TRUE}
# Get local directory and load the file

localDir <- getwd()
chessFile <- paste(str_trim(localDir),"//tournamentinfo.txt",sep="")

# Skip top 3 lines as it contains header info

chess.text<-read_lines(chessFile,skip = 3)
head(chess.text,10)
```

There are total of 64 players. File needs formatting in order to perform calculations and generate CSV output.

```{r , echo=TRUE}
# First level of formatting removes hypen("-") characters

chess.text.level1 <- gsub("-","",chess.text)
head(chess.text.level1, 6)
```

```{r , echo=TRUE}
# Second level of formatting removes empty lines
# This makes a continous data without newline characters 

chess.text.level2 <- paste(unlist(chess.text.level1), collapse = " ")
head(chess.text.level2,10)
```

```{r , echo=TRUE}
# Add extra characters at the end of text data. This helps read complete data. 
# If extra characters are not added, last line will be skipped. 

chess.text.level2 <- paste(chess.text.level2,"_*_*_*_")

# Extract the text using pattern repeating 180 characters
# Idea is to combine 2 lines of each player's info into single line
# Once combined, information about single player takes 180 characters

pattern <- "[[:print:]]{180}"
chess.text.level3 <- unlist(str_extract_all(chess.text.level2, pattern = pattern))
head(chess.text.level3,3)
```

```{r , echo=TRUE}
# Each player has about 20 columns of information. Split the single string into 20 columns

chess.data <- data.frame(unlist(str_split_fixed(chess.text.level3, "\\|", 21)))
head(chess.data, 5)
```

```{r , echo=TRUE}
# Using regular expressions extract the pre match rating for each player

pattern = "(R:\\s+[[:digit:]]{3,4})"
chess.preGameRating <- unlist(str_extract(string = chess.data[,12], pattern = pattern))

# Get the pre-game rating
pattern = "[[:digit:]]{3,4}"
chess.preGameRating <- str_extract(string = chess.preGameRating, pattern = pattern)
chess.preGameRating <- data.frame(chess.data, preGameRating = chess.preGameRating)

head(chess.preGameRating, 5)

```

```{r , echo=TRUE}
# Format the column type

chess.preGameRating$X1 <- as.numeric(as.character(chess.preGameRating$X1))
chess.preGameRating$preGameRating <- as.numeric(as.character(chess.preGameRating$preGameRating))

```

```{r , echo=TRUE}
# Extract the information about opponent for each round
# Opponent player info is set to "0" for all "NA" values

pattern = "[[:digit:]]{1,2}"

chess.preGameRating$G1opponent = ifelse(is.na(str_match(string = chess.preGameRating$X4, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X4, pattern = pattern)))
chess.preGameRating$G2opponent = ifelse(is.na(str_match(string = chess.preGameRating$X5, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X5, pattern = pattern)))
chess.preGameRating$G3opponent = ifelse(is.na(str_match(string = chess.preGameRating$X6, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X6, pattern = pattern)))
chess.preGameRating$G4opponent = ifelse(is.na(str_match(string = chess.preGameRating$X7, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X7, pattern = pattern)))
chess.preGameRating$G5opponent = ifelse(is.na(str_match(string = chess.preGameRating$X8, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X8, pattern = pattern)))
chess.preGameRating$G6opponent = ifelse(is.na(str_match(string = chess.preGameRating$X9, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X9, pattern = pattern)))
chess.preGameRating$G7opponent = ifelse(is.na(str_match(string = chess.preGameRating$X10, pattern = pattern)), 0, as.numeric(str_extract(string = chess.preGameRating$X10, pattern = pattern)))

head(chess.preGameRating,10)
```

```{r, echo=TRUE}
chessTable <- data.table(chess.preGameRating)

chess.rating <- sqldf("SELECT ct1.*,SUM(ct2.preGameRating) AS sumOfOpponentPreGameRating,
                          SUM(CASE WHEN ct1.G1opponent > 0 THEN 1 
                          WHEN ct1.G2opponent > 0 THEN 1
                          WHEN ct1.G3opponent > 0 THEN 1
                          WHEN ct1.G4opponent > 0 THEN 1
                          WHEN ct1.G5opponent > 0 THEN 1
                          WHEN ct1.G6opponent > 0 THEN 1
                          WHEN ct1.G7opponent > 0 THEN 1
                          ELSE 0 END) AS numberOfGamesPlayed
                      FROM chessTable ct1, chessTable ct2 
                      WHERE ct2.X1 IN(ct1.G1opponent,ct1.G2opponent,ct1.G3opponent,ct1.G4opponent,ct1.G5opponent,ct1.G6opponent,ct1.G7opponent) GROUP BY ct1.X1")

head(chess.rating,5)
```

```{r, echo=TRUE}
# Average opponent pre game rating (rounded).  

chess.rating$avgOppPreGameRating <- round(chess.rating$sumOfOpponentPreGameRating / chess.rating$numberOfGamesPlayed,0)
head(chess.rating,10)
```

```{r, echo=TRUE}
# Change column names to readable format

names(chess.rating)[names(chess.rating) == "X1"] <- "playerNumber"
names(chess.rating)[names(chess.rating) == "X2"] <- "playerName"
names(chess.rating)[names(chess.rating) == "X11"] <- "playerState"
names(chess.rating)[names(chess.rating) == "X3"] <- "playerPoints"
names(chess.rating)[names(chess.rating) == "preGameRating"] <- "playerPreGameRating"


# Format the column values remove extra spaces

trim <- function (x) gsub("^\\s+|\\s+$", "", x)
chess.rating$playerState <- trim(chess.rating$playerState)
chess.rating$playerPoints <- trim(chess.rating$playerPoints)
chess.rating$playerName <- trim(chess.rating$playerName)

# Create subset of data to be written to CSV file

chess.csvdata <- subset(chess.rating,select = c(playerNumber,playerName,playerState,playerPoints,playerPreGameRating,avgOppPreGameRating))

head(chess.csvdata,10)
```

```{r, echo=TRUE}
# Generate CSV files

chessOutFile <- paste(str_trim(localDir),"//chessOutFile.csv",sep="")
write.table(chess.csvdata, file = chessOutFile, row.names = FALSE, sep = ",")
```



References: 

1. SQL from within R - https://github.com/ggrothendieck/sqldf#example-4-join

2. Trim function - http://stackoverflow.com/questions/2261079/how-to-trim-leading-and-trailing-whitespace-in-r

3. Regular expressions - Handling and Processing Strings in R, Gaston Sanchez

